---
title: "outbreakteachR package tutorial overview"
author: "OJ Watson"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: show
    css: knitr.css
    fig_caption: yes
    theme: readable
    toc: yes
    toc_float: yes
---

## Overview 

  The outbreakteachR R package has been designed to be an additional tool in helping assist
with running the "paper outbreak" practical. The package assumes that the data has
been collected and stored in a similar fashion to that presented within the xlsx file
included within the external data directory, and that the main data from the first 
sheet in the excel document, between cell A2:Oxxx, is to be used.

  Given the above specifications, the tutorial below will give an overview of how
to read in the data and produce a number of plots to help visualise the outbreak and
present the key epidemiological parameters.

## Loading the data

First we will read in the data from wherever you have stored the .xlsx file. Below
we will be using the example included with the package. The output of the following
section is stored within the package as well as a sample dataset called "paper.outbreak.2016".
If you are using the sample dataset in the remainder of the tutorial then remember to
set the outbreak.dataset variable equal to paper.outbreak.2016.


```{r Loading the data, include=TRUE, message = FALSE, warning = FALSE}

# First make sure the package is installed
#devtools::install_github("OJWatson/outbreakteachR")

# Load package - Warning messages will be thrown however these are not a problem
library(outbreakteachR)

# With the package installed first read in the data.
# This function will throw warnings highlighting rows where there may be errors in the excel file
# Remember to replace the argument for xlsx.file to the local file path where your saved worksheet is
# If it is on your desktop it will look something like xlsx.file = "C:/Users/Bob/Desktop/savedworksheet.xlsx"
outbreak.dataset <- outbreakteachR::outbreak_dataset_read(xlsx.file = system.file("extdata","2016_solutions_final.xlsx",package="outbreakteachR"),
                                                 attempt.imputation = TRUE)

# Visualise the dataset
str(outbreak.dataset)

# This dataset is also saved within the outbreakteachR package and can be accessed as such:
data("paper.outbreak.2016",package="outbreakteachR")
## Confirm the datasets are the same
identical(paper.outbreak.2016,outbreak.dataset)

# If you are using the sample dataset for the remainder of the tutorial then 
# uncomment the following line of code:

# outbreak.dataset <- paper.outbreak.2016
```

## Create an infection network

For the next task we will need to subset the collected data so that we are only
looking at those cases that were not reinfections. Using this subset of the data
we will then be able to plot a number of static images from the network that will
enable us to sere how the outbreak progressed in time.

```{r Create an infection network, include=TRUE, fig.align = "center", cache = T}

# Convert the read outbreak dataset into only the non-reinfections
first_infection_list <- outbreakteachR::first_infection_list(outbreak.dataset)

# The above function has now sorted our data into a list of two dataframes. These
# data frames show the "linelist" and "contacts" data. We can now see that the
# linelist data only possesses information about contacts that were not reinfections:

unique(first_infection_list$linelist$Reinfection)

# Next we will create a plot of the outbreak that is time-orientated, i.e. the relative
# node positions represents the point in time that an individual was infected:

outbreakteachR::infection_network_plot(first_infection_list = first_infection_list, time = TRUE, log = FALSE)

# We might find that the outbreak is quite cramped in teh above as a lot of infections happen over 
# a very short time frame with respect to the 5 day outbreak. We can thus log transform the times of
# infection to spread out the outbreak for further clarity at the earlier stages

outbreakteachR::infection_network_plot(first_infection_list = first_infection_list, time = TRUE, log = TRUE)

# We can also visualise the network in a more "tree" like fashion which shows each
# infection event with an equal length branch for easier visualisation:

outbreakteachR::infection_network_plot(first_infection_list = first_infection_list, time = FALSE)

# Lastly we can visualise what the outbreak looked like by viewing the outbreak
# at the end of each day, i.e. at 24 hour intervals

outbreakteachR::daily_timeseries_plot(first_infection_list=first_infection_list)

```

## Animate the infection network

So far we have only been able to visualise the outbreak at discrete moments in
time. To improve this we can also create an html page showing an animation of the
outbreak process. (This will take a minute or two).

```{r Animate the infection network, include=TRUE, message = FALSE, warning = FALSE, fig.align = "center", cache=TRUE}

# Change the file parameter to wherever you want the html to be stored
  outbreakteachR::animate_infection_network(first_infection_list=first_infection_list,
                            file=paste(getwd(),"/Animated-Network-Dynamic.html",sep=""),
                            year=2016,
                            detach=FALSE)
                            
```

When you run the above function the animated network will automatically load in
a web browser and save the html page to file path specified. This html is embedded
below, and the animated network can be viewed by clicking play, and the duration
of the animation changed with the menu in the top left. To view the html page
output in a new tab please click [here](https://cdn.rawgit.com/OJWatson/outbreakteachR/master/tutorials/Animated-Network-Dynamic.html)


<iframe width="800" height="630" frameBorder="0" src="https://cdn.rawgit.com/OJWatson/outbreakteachR/master/tutorials/Animated-Network-Dynamic.html" allowfullscreen></iframe>

## Epidemic Time Series

We can also use the collected data to plot the epidemic time series, using the times
recorded for when the infection started and ended in each individual.

```{r Epidemic Time Series, include=TRUE, message = FALSE, warning = FALSE, fig.align = "center", cache = TRUE}

outbreakteachR::epidemic_timeseries_plot(outbreak.dataset = outbreak.dataset)

```

## Parameter investigation

We can also examine the distribution of the epidemiological parameters.

```{r Epidemiological parameter investigation, include=TRUE, message = FALSE, warning = FALSE, fig.align = "center"}

outbreakteachR::paramater_boxplots_plot(outbreak.dataset=outbreak.dataset)

```

We might also want to view the offspring distribution to see how it compares to the Poisson distribution that was used 
to generate the actual observed data. 

```{r Offspring distribution, include=TRUE, message = FALSE, warning = FALSE, fig.align = "center"}

outbreakteachR::offspring_distribution_plot(outbreak.dataset=outbreak.dataset,include.reinfectons = TRUE,title = 2016)

```

What we can see from above is the confirmed number of secondary cases from each individual,
which has a mean very close to 1.8. However if we might also want to look at the distribution
describing only the successful contacts, i.e. those that led to infections rather than reinfections. 

```{r Offspring distribution with zero, include=TRUE, message = FALSE, warning = FALSE, fig.align = "center"}

outbreakteachR::offspring_distribution_plot(outbreak.dataset=outbreak.dataset, include.reinfectons = FALSE,title=2016)

```

## Simulated outbreaks

It can often be very useful to see how this outbreak may compare to a simulated outbreak.
To do this we can use the following function to simulate 200 outbreaks and plot
the interquartile range of these simulations: 

```{r Simulated outbreaks, include=TRUE, message = FALSE, warning = FALSE, fig.align = "center"}

outbreakteachR::bootstrap_simulated_plot(
  first_infection_list = first_infection_list,
  outbreak.dataset = outbreak.dataset, 
  lower.quantile=0.25,
  upper.quantile=0.75,
  include.observed = T,
  replicates = 200)

```

To give some more information into how the simulated outbreaks are modeled. 
For these simulations we assume each simulation is seeded with three infections 
at the same time as the observed outbreak, and has a total population size equal to the 
outbreak dataset. The number of secondary infections is then drawn from a Poisson 
distribution with a mean of 1.8, as was used in the practical. The time of each 
infection  is also drawn from a Poisson with a mean equal to the observed 
mean generation times. The infected individual then recovers after a period of time, 
which is sampled from a Poisson with mean equal to the mean infectious period. 
If they do not infect anyone then they recover after a period of time drawn from a 
Poisson with mean equal to the mean recovery time for the observed outbreak. 
Simulations are then stopped after a period of time equal to the length of the observed outbreak.

Given this simulation methodology we can start to hypothesise about why the simulated
outbreaks are delayed when compared to the observed. If we look back at the box plots
of the generation and recover times we can see that the data clearly does not 
follow a Poisson distribution, but is multimodal with a period of approximately 
2 hours. As a result we might choose to conduct the simulations again but sampling
generation times, infectious periods and recovery times from the observed outbreak:

```{r Simulated outbreaks with sampling, include=TRUE, message = FALSE, warning = FALSE, fig.align = "center"}

outbreakteachR::bootstrap_simulated_plot(
  first_infection_list = first_infection_list,
  outbreak.dataset = outbreak.dataset, 
  lower.quantile = 0.25,
  upper.quantile = 0.75,
  include.observed = T,
  replicates = 200,
  sampling = TRUE)

```

This simulation is much better, with the timing of the infection peak much closer
to the observed data. This sampling method however still fails to perfectly capture the
temporal dynamics exhibited. There are many reasons why this is, but we can begin
to understand why this is by simply seeing how many infections there are that occurred on
day 1, which is much larger than would be expected given 3 seeds and the generation
times that we sampled. 

## Summary and further thoughts

Hopefully the above tutorial has shown how the *outbreakteachR* outbreak practical can be extended
using R to extend the analysis and aid in visualising the outbreak process.

---


